Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-27 application.tex
writing summary now

\section{Application to protoplanetary disks}\label{application} 
As an application of our results, we estimate where in a 
protoplanetary disk do the thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental isothermal VSI, 
for which we can use the simple criterion Eq. \ref{iso_vsi_cond}. The
condition derived below may be considered as a sufficient condition for the VSI.

In our disk models we have adopted a simple thermal relaxation
function in the energy equation. To connect our model with a realistic
disk, we consider the more physically-motivated form of the energy
equation with radiative cooling 
\begin{align}\label{real_energy}
\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where the temperature $T$ is defined such that $P=\mathcal{R}\rho
T/\mu$, where $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight. In Eq. \ref{real_energy} $S$ is the
specific entropy,
\begin{align}
  S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
\end{align}
where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the specific heat capcity at
constant volume, $D/Dt$ is the convective derivative, $\bm{F}$ is the
radiative flux, and $Q_+$ is a heating term such that an equilibrium
solution exists.

Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
the thermal relaxation time $t_c$ as used in our linear
calculations. To do so, we first consider separately perturbations
with radial lengthscales $L\equiv 2\pi/k_x\ll l_\mathrm{rad}$ and 
$L\gg l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

@@

\subsection{Radiative diffusion}
For perturbation lengthscales $L\gg l_\mathrm{rad}$, let us assume that
temperature fluctuations are smoothed out by radiative diffusion. To
model this in linear theory, we choose the heating term such that
\begin{align}
  Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
    \left(T-T_{t=0}\right)\right], 
\end{align} 
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly. Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
    c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right). 
\end{align}
This, of course, fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates. We thus proceed by assuming vertical
derivatives can be neglected in Eq. \ref{diff_cool_proper} and write
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
%we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}. 
\end{align}
From Eq. \ref{diff_cool_approx} we identify the thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the 
dimensionless diffusion coefficient defined for convenience, and is
approximately 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \simeq  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}

@@

The dimensionless thermal relaxation time $\beta$ is then
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H_\mathrm{iso}^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right).
\end{align}

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).  
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model, 
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7},
\end{align}
and we recall $\beta < \beta_\mathrm{crit}$ permits the fundamental 
isothermal VSI to operate.

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the MMSN with $\mu = 2.33$ and
$\gamma=1.4$. Although it is possible to satisfy
$\beta<\beta_\mathrm{crit}$ at small radii by choosing a sufficiently
large $\khat$, our linear calculations suggest growth rates eventually
decay as $\khat\to\infty$. We thus restrict our attention to
$\khat\lesssim 10^2$.

Fig. \ref{mmsn_bcrit_bcool} suggests that the fundamental isothermal
VSI can operate in the outer parts of the MMSN at radii on the order
of tens of AU. However, the thermal relaxation time $\beta$ is not
much shorter than $\beta_\mathrm{crit}$, so the VSI may be
weak.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_mmsn}  
  \caption{Dimensionless thermal relaxation timescales $\beta$ in the MMSN
    using Eq. \ref{beta_mmsn} for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot). The relaxation timescale
    $\beta_\mathrm{crit}$, below which the fundamental isothermal VSI
    operates, is shown as the solid line. 
    \label{mmsn_bcrit_bcool}}   
\end{figure}

\subsection{Numerical example}
Consider  the MMSN at $r=50\mathrm{AU}$, with $\gamma=1.4$ and
$\mu=2.33$. Eq. \ref{mmsn_epsilon} gives $\epsilon \simeq
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The disk supports the fundamental isothermal VSI, but only
marginally. Using the above value of $\beta$, MMSN disk parameters and
$\khat=10$ in our linear code, we find a 
growth rate of 
\begin{align*}
  \nu = 4.0\times10^{-3}\Omega_k.
\end{align*}
This corresponds to a characteristic growth time $\sim 40$
orbits. 
%This mode was found to be the most unstable for
%$\khat=10$.

% and
% physically acceptable -> other modes
% are crazy

%We also solved the linear problem with varying $\khat$  
%and $\beta$ determined by Eq. \ref{beta_mmsn}. We find the most
%unstable wavenumber occurs at $\khat\simeq60$ with $\nu\simeq
%3.6\times10^{-3}\Omega_k$, i.e. only slightly faster than $\khat=20$.   
%However, the eigenfunctions for large $\khat$ were found to have large
%amplitude oscillations near the vertical boundaries. This is shown in
%Fig. \ref{mmsn_eigenW}. These modes may not be important since they
%mostly affect the disk atmosphere where there is little mass.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/eigenvectorW_mmsnkx10.ps}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0cm]{figures/eigenvectorW_mmsnkx60.ps}
  \caption{Most unstable VSI modes in the MMSN at
    $r=30\mathrm{AU}$. The eigenfunction $W$ for $\khat=20$ (top) and
    $\khat=60$ (bottom) are shown. The real (imaginary) part of $W$
    is plotted as the solid (dotted) lines.\label{mmsn_eigenW}}  
\end{figure}

% only find one unstable mode -> close to marginal 
% larger kx

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our results, we estimate where in a 
protoplanetary disk do thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental isothermal VSI, 
for which we can use the criterion Eq. \ref{iso_vsi_cond}.

In our disk models we have adopted a simple thermal relaxation 
function in the energy equation. To connect our model with a realistic
disk, we consider a more physically-motivated form of the energy
equation with radiative cooling 
\begin{align}\label{real_energy}
\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, where $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight. In Eq. \ref{real_energy} $S$ is the
specific entropy,
\begin{align}
  S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
\end{align}
where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
specific heat capcity at constant volume, $D/Dt$ is the convective
derivative, $\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists when $D/Dt=0$.

Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
the thermal relaxation time $t_c$ as used in our linear
calculations. To do so, we first consider separately perturbations
with radial lengthscales $L\equiv 2\pi/k_x\ll l_\mathrm{rad}$ and 
$L\gg l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity. 
%need to comment on kappa_d(rho,T)

@@

\subsection{Radiative diffusion}
For perturbation lengthscales $L\gg l_\mathrm{rad}$, let us assume that
temperature fluctuations are smoothed out by radiative diffusion. To
model this in linear theory, we choose the heating term such that
\begin{align}
  Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
    \left(T-T_{t=0}\right)\right], 
\end{align} 
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly. Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
    c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right). 
\end{align}
This, of course, fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed
by assuming vertical derivatives can be neglected in
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}. 
\end{align}
From Eq. \ref{diff_cool_approx} we identify the thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the 
dimensionless diffusion coefficient.

@@

In order to utilize our linear results, which assumes $t_c$ is a
constant, we need to make further 
approximations. Specifically, we consider vertically-isothermal disks
and replace explicit $\rho$ depedences in the above expressions by $\Sigma/
H_\mathrm{iso}$, where $\Sigma$ is the surface density at the fiducial
radius of interest. For example, the dimensionless diffusion
coefficient becomes 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
Furthermore, we will choose a temperature-dependent opacity model for
simplicity, so that $\kappa_d=\kappa_d(T)$. The dimensionless thermal
relaxation time $\beta$  
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H_\mathrm{iso}^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right)
\end{align}
is then a constant for vertically-isothermal disks, 
and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is a neccessary condition for the
fundamental isothermal VSI when $\khat\gg 1$. For $\khat\to0$, this
becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the Minimum Mass Solar Nebula (MMSN) by setting
the scalings $\hat{T}=\hat{\Sigma}=\hat{\kappa}_d=1$, with $\mu =
2.33$ and $\gamma=1.4$. The figure suggests that the fundamental
isothermal VSI operates in the outer parts of the MMSN at radii on
the order of tens of AU. However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak.

We remark that it is possible to satisfy $\beta<\beta_\mathrm{crit}$
at small radii by choosing a sufficiently large $\khat$. However, our
linear calculations suggest growth rates eventually decay as 
$\khat\to\infty$. Thus, very small-scale VSI in the inner disk is
probably unimportant.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_mmsn}  
  \caption{Dimensionless thermal relaxation timescales $\beta$ in the MMSN
    using Eq. \ref{beta_mmsn} for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot). The solid line is the
    critical thermal relaxation timescale $\beta_\mathrm{crit}$, and 
    $\beta<\beta_\mathrm{crit}$
    is necessary for the fundamental isothermal VSI to operate  
    for $\khat\gg1$. 
    \label{mmsn_bcrit_bcool}}   
\end{figure}

\subsection{Numerical example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The disk is unstable to the fundamental isothermal VSI, but not
strongly.

Using the MMSN disk parameters at $r=50\mathrm{AU}$ and the
scale-dependent thermal relaxation $\beta(\khat)$ given by
Eq. \ref{beta_mmsn} in our linear code, we calculate the fundamental
VSI growth rates in Fig. \ref{mmsn_growth}. Growth rates
are weakly depedent on $\khat$ until $\khat\gtrsim50$ beyond which the
mode is rapidly stabilized. A representative growth rate is 
\begin{align*}
  \nu =0.059\epsilon\Omega_k = 4.0\times10^{-3}\Omega_k 
\end{align*}
for $\khat=10$. This corresponds to a characteristic growth time $\sim
40$ orbits.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_mmsn_50AU} 
  \caption{VSI growth rates in the MMSN at
    $r=50\mathrm{AU}$. The thermal relaxation timescale $\beta(\khat)$
    used for this calculation is given by
    Eq. \ref{beta_mmsn} and is scale-dependent. 
    \label{mmsn_growth}}  
\end{figure}

Fig. \ref{mmsn_eigenW} show several eigenfunctions for the modes in
Fig. \ref{mmsn_growth}. As $\khat$ is increased, the disturbance
become more concentrated near the vertical boundaries. Thus, modes
with very large $\khat$ may not be important in practice because they
predominantly affect the disk atmosphere, which contains little mass.

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx1.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx10.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx50.ps} 
  \caption{Fundamental VSI modes in the MMSN at
    $r=50\mathrm{AU}$. The eigenfunction $W$ for $\khat=1$ (top),
    $\khat=10$ 
    (middle) and $\khat=50$ (bottom) are shown. The real (imaginary)
    part of $W$ is plotted as the solid (dotted)
    lines. The thermal relaxation timescale $\beta(\khat)$ is given by
    Eq. \ref{beta_mmsn} and is scale-dependent. \label{mmsn_eigenW}}    
\end{figure}
% only find one unstable mode -> close to marginal 
% larger kx

### 2015-01-27 linear_adia.tex
writing summary now

\subsubsection{Lower limit on the thermal relaxation timescale for the
  fundamental adiabatic VSI mode}
{\bf Note: for reference only. Not to be included in paper.} 
Fig. \ref{relax_bcrit} indicates there exists a lower limit on the
thermal relaxation time for the system to behave adiabatically and
support the fundamental adiabatic VSI. This critical
timescale satisfies Eq. \ref{relax_cond1}---\ref{relax_cond2} with 
\begin{align}
  \beta \sim \beta_0\khat^{1/2},\quad \hat{\omega}  \sim
  \hat{\omega}_0\khat^{-1/2} \text{ as } \khat\to\infty.
\end{align}
The constants $\beta_0$ and $\hat{\omega}_0$ are determined by the
simultaneous equations
\begin{align}
  & 0 = \gamma\beta_0^2(1-\gamma) + 2\epsilon q \gamma
  \beta_0\hat{\omega}_0 - \hat{\omega}_0^2 +
  \beta_0^2\gamma^2\hat{\omega}_0^4.\\
  & 0= \beta_0(1-\gamma) + \epsilon q \hat{\omega}_0 - \epsilon q
  \gamma^2 \beta_0^2\hat{\omega}_0^3. 
\end{align}
For $\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$, the appropriate
solution is $\beta_0 = 10.33$ and $\hat{\omega}_0 = 0.74$.

>>>>

% \subsubsection{Lower limit on the thermal relaxation timescale for the
%   fundamental adiabatic VSI mode}
% {\bf Note: for reference only. Not to be included in paper.} 
% Fig. \ref{relax_bcrit} indicates there exists a lower limit on the
% thermal relaxation time for the system to behave adiabatically and
% support the fundamental adiabatic VSI. This critical
% timescale satisfies Eq. \ref{relax_cond1}---\ref{relax_cond2} with 
% \begin{align}
%   \beta \sim \beta_0\khat^{1/2},\quad \hat{\omega}  \sim
%   \hat{\omega}_0\khat^{-1/2} \text{ as } \khat\to\infty.
% \end{align}
% The constants $\beta_0$ and $\hat{\omega}_0$ are determined by the
% simultaneous equations
% \begin{align}
%   & 0 = \gamma\beta_0^2(1-\gamma) + 2\epsilon q \gamma
%   \beta_0\hat{\omega}_0 - \hat{\omega}_0^2 +
%   \beta_0^2\gamma^2\hat{\omega}_0^4.\\
%   & 0= \beta_0(1-\gamma) + \epsilon q \hat{\omega}_0 - \epsilon q
%   \gamma^2 \beta_0^2\hat{\omega}_0^3. 
% \end{align}
% For $\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$, the appropriate
% solution is $\beta_0 = 10.33$ and $\hat{\omega}_0 = 0.74$.

### 2015-01-27 numerical.tex
writing summary now

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_imag_bloop}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag_bloop2}
  \caption{Growth rates of the fundamental VSI mode in 
    nearly-vertically isothermal disks ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$ and a range of thermal relaxation timescales 
    $\beta$. The eigenfrequencies are calculated by numerically 
    solving the full linear eigenvalue problem. \label{relax_growth_num}}   
\end{figure}

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.125$. The latter value is close to the critical
value beyond which the fundamental isothermal VSI is stabilized
(\S\ref{iso_vsi_beta_crit}).  Increasing $\beta$ restricts 
the region in which $W\sim z$ closer to the midplane, with oscillatory
behavior emerging at the boundaries, where buoyancy first becomes
important.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/eigenvectorW_beta0d01}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0cm]{figures/eigenvectorW_beta0d125}
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a disk with dimensionless thermal relaxation timescale
    $\beta \equiv t_c\Omega_k = 0.01$ (top) and $0.125$ (bottom) for
    perturbations with a radial wavenumber $k_xH_\mathrm{iso}=10$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line. Other disk parameters are
    $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.
    \label{relax_eigenW_num}}  
\end{figure}

@@

\subsection{Vertically non-isothermal disks}
We briefly consider vertically non-isothermal disks with
$\Gamma=1.4$. Other disk parameters are $p=0$, $q=-1.0$ and $\epsilon
= 0.05$. These are again chosen for comparison with \cite{nelson13}. 
In this case we set $\zmax\simeq2.2\epsilon r$, which is close to the
zero-density surface.

\subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}

Consider first the case $\gamma=1.4$ so that there is no vertical
entropy gradient. We plot in Fig. the growth rates for the
fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
do not reach zero, implying the VSI operates regardless of the thermal
relaxation timescale if $\gamma=\Gamma$. This result is consistent
with \cite{nelson13}.

We also note that growth rates for $\beta = 10^{-2}$ and
$\beta=10^{2}$ behave similarly. This is not surprising because
the linearized energy equation (Eq. \ref{lin_energy}) implies that 
\begin{align}
  W = \frac{\left(\Gamma^{-1} -
      \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
  Q \text{ for } \gamma\equiv \Gamma.   
\end{align}
Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
$\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
to produce qualitatively similar results.

@@

Interestingly, Fig. \ref{growth_vnoniso1} show that growth rates
maximize for $\beta=O(1)$. However, all growth rates are similar to
order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

\subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}

%gamma = 2.0, Gamma = 1.4, vary bcool

Finally, we introduce buoyancy in a vertically non-isothermal disk by
setting $\gamma=2$. This situation was not simulated in 
\cite{nelson13}, but is similar in spirit with vertically isothermal 
disks evolved with $\gamma>1$, as considered in the previous section.

Fig. plots the growth rates for this case. The trend is
qualitatively similar to that for vertically isothermal disks with
$N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth rates reach
zero at finite $\khat$.

>>>>

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
   0cm]{figures/compare_eigen_imag_bloop} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/compare_eigen_imag_bloop2} 
  \caption{Growth rates of the fundamental VSI mode in 
    nearly-vertically isothermal disks ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ under a
    range of thermal relaxation timescales   
    $\beta$. The eigenfrequencies are calculated by numerically 
    solving the full linear eigenvalue problem. \label{relax_growth_num}}   
\end{figure}

We visualize the effect of thermal relaxation in 
Fig. \ref{relax_eigenW_num} by comparing the eigenfunction $W$ for
$\beta=0.01$ and $\beta=0.125$. The latter value is close to the critical
value beyond which the fundamental isothermal VSI is stabilized
(\S\ref{iso_vsi_beta_crit}). Hence the latter growth rate is smaller
than the former by a factor $\sim 20$. Increasing $\beta$ restricts 
the region in which $W\sim z$ closer to the midplane, with oscillatory
behavior emerging at the boundaries, where buoyancy first becomes
important. % We also find this trend for fixed $\beta$ but increasing
% $\khat$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_beta0d125} 
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta \equiv t_c\Omega_k = 0.01$
    (top) and $0.125$ (bottom). The
    perturbation radial wavenumber is $\khat_\mathrm{iso}=10$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line.
    \label{relax_eigenW_num}}  
\end{figure}

@@

\subsection{Vertically non-isothermal disks}
We briefly consider vertically non-isothermal disks with
$\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
\cite{nelson13}. In this case we set $\zmax\simeq2.2\epsilon r$, which
is close to the zero-density surface.

\subsubsection{$\gamma=\Gamma \neq 1$, $N_z^2\equiv 0$}
Consider first the case $\gamma=1.4$ so that there is no vertical
entropy gradient. We plot in Fig. the growth rates for the
fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
do not reach zero, implying the VSI operates regardless of the thermal
relaxation timescale if $\gamma=\Gamma$. This result is consistent
with \cite{nelson13}.

We also note that growth rates for $\beta = 10^{-2}$ and
$\beta=10^{2}$ behave similarly. This is not surprising because
the linearized energy equation (Eq. \ref{lin_energy}) implies that 
\begin{align}
  W = \frac{\left(\Gamma^{-1} -
      \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
  Q \text{ for } \gamma\equiv \Gamma.   
\end{align}
Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
$\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
to produce qualitatively similar results. Interestingly,
Fig. \ref{growth_vnoniso1} show that growth rates 
maximize for $\beta=O(1)$. However, all growth rates are similar to
order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

@@

\subsubsection{$\gamma\neq\Gamma \neq 1$, $N_z^2\geq0$}
%gamma = 2.0, Gamma = 1.4, vary bcool 
Finally, we introduce buoyancy in a vertically non-isothermal disk by
setting $\gamma=2$. This scenario was not considered by  
\cite{nelson13}, but is similar in spirit with vertically isothermal 
disks evolved with $\gamma>1$, as considered in the previous section. 
Fig. \ref{growth_vnoniso2} plots the fundamental VSI growth rates for
this case. We observe the same trend as for vertically isothermal
disks with $N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth
rates reach zero at finite $\khat$.
